---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# linktree

<!-- badges: start -->
<!-- badges: end -->

The goal of `linktree` is to estimate the transmission assorativity coefficient.

## Installation

You can install the development version of linktree from [GitHub](https://github.com/CyGei/linktree/) with:

``` r
# install.packages("devtools")
devtools::install_github("CyGei/linktree")
```

## Understanding the transmission assortativity coefficient
`linktree` leverages transmission chain data to estimate group transmission assortativity; this quantifies the extent to which individuals transmit within their own group compared to others.

### Gamma
$\gamma$ is defined as the excess probability of a secondary infection taking place within-group compared to random expectation. $\gamma$ values range from 0 (fully disassortative) to $\infty$ (fully assortative), with 1 indicating homogeneous patterns. For instance, $\gamma_a = 2$ indicates that an infected individual from group $a$ is twice as likely to infect an individual from the same group compared to infecting an individual from another group.

### Delta
To simplify interpretation, we introduce a rescaled parameter $\delta$, ranging between -1 (fully disassortative) and 1 (fully assortative), with 0 indicating homogeneous patterns. 

We can visualise the relationship between $\gamma$ and $\delta$:

```{r}
library(linktree)
delta <- seq(-1, 1, 0.01)
gamma <- delta2gamma(delta)
plot(gamma, delta, type = "l")
```

### Estimation
`get_gamma` and `get_delta` estimate $\gamma$ and $\delta$ from a transmission tree. It requires the following inputs:
- `from`: a vector of infector's group (e.g. age group, sex, vaccination status, etc.)
- `to`: a vector of infectee's group
- `f`: a vector of the groups' sizes or relative sizes (e.g. proportion of the population in each group).

## Example

### Data
We can load the `simulated_tree` data from the package. It represents a simulated transmission tree in a population with 2 groups (A & B). The tree is stored as a data frame with 4 columns: `id`, `group`, `date_infection` and `date_onset`. The `id` column contains the unique identifier of each individual. The `group` column contains the group to which each individual belongs. The `date_infection` column contains the date of infection of each individual. The `date_onset` column contains the date of onset of symptoms of each individual. `source` and `source_group` are the unique identifier of the individual who infected the individual and the group to which the source individual belongs.

Group A is assortative with a `delta` coefficient of 0.75 and group B is neutral with a `delta` coefficient of 0. Both group sizes were set at 100 individuals.
This outbreak was simulated using the package [o2groups](https://github.com/CyGei/o2groups).

```{r}
head(simulated_tree)
```


### Visualise the transmission tree & outbreak

```{r}
#devtools::install_github("reconhub/epicontacts", ref = "timeline")
library(epicontacts)
library(dplyr)
library(ggplot2)


ttree<- epicontacts::make_epicontacts(
  linelist = select(simulated_tree, c(group, id, date_infection, date_onset)),
  contacts = select(simulated_tree, c(source, id, source_group, group)),
  id = "id",
  from = "source",
  to = "id",
  directed = TRUE
)

plot(
  ttree,
  selector = FALSE,
  x_axis = "date_onset",
  method = "ggplot",
  node_color = "group",
  edge_color = "source_group"
)+
  scale_colour_manual(values = c("orange", "purple"))+
  scale_fill_manual(values = c("orange", "purple"))+
  labs(title = "Transmission tree", x = "Date of onset", y = "")


#Plot the epidemic curve
simulated_tree %>%
  ggplot(aes(x = date_onset)) +
  geom_bar(stat = "count", aes(fill = group)) +
  scale_fill_manual(values = c("orange", "purple")) +
  labs(title = "Epidemic curve", x = "Date of onset", y = "Number of cases")

```


### Estimate the transmission assortativity coefficient

```{r}
delta <- get_delta(
  from = simulated_tree$source_group,
  to = simulated_tree$group,
  f = c(A = 100, B = 100)
)

delta
```
Group-level saturation (i.e. the depletion of susceptibles) can significantly impact the estimation of the transmission assortativity coefficient. To account for this, we should estimate the transmission assortativity coefficient before the epidemic peak (i.e. before day 17).

```{r}
cut_tree <- simulated_tree %>%
  filter(date_onset <= 17)

delta <- get_delta(
  from = cut_tree$source_group,
  to = cut_tree$group,
  f = c(A = 100, B = 100)
)
plot(delta)
```
